---
title: "KBL_stats"
author: "Arlin Stoltzfus"
date: "June 11, 2015"
output: html_document
---

The observed results: 
```{r hardcoded_kbl_results}
obs_all <- 457  # all duplicate pairs
obs_accel <- 76 # pairs with at least 1 member accelerated
obs_both <- 4   # pairs with both members accelerated
```

## Null models 

### Uncorrelated

Calculate the expected values 
both accelerated: p2 = 0.07661, expect 3.501 pairs, observed 4 pairs
one accelerated: 2 * p * (1 - p) = 0.1597, expect 73.00 pairs, observed 72 pairs 
none accelerated: (1 - p)2 = 0.8326, expect pairs 380.5 pairs, observed 381 pairs 
There are various ways to calculate a P value for this.  The simplest is just a chi-squared test on 3 categories with 2 degrees of freedom.  
```{r chi_squared_test}
p <- 80/914
exp_both <- p^2 
exp_one <- 2 * p * (1 - p) 
exp_none <- (1 - p)^2 
obs <- c(obs_both, obs_accel - obs_both, obs_all - obs_accel) 
exp <- c(exp_both, exp_one, exp_none)
  
chisq.test(x = obs, p = exp)
```

Randomly assigned rates for pairs of genes (a1, b1), (a2, b2) and so on up to `r obs_all`
```{r random_independent_pairs}
r_ave <- 100  # rate of evolution, arbitrary units
r_sd <- 10    # standard deviation of rate among pairs
dupls <- data.frame(a = rnorm(obs_all, r_ave, r_sd), b = rnorm(obs_all, r_ave, r_sd))
```

Now plot out the distribution, showing which pairs are both accelerated (upper right quadrant), singly accelerated (upper left, lower right), or unaccelerated (lower left). 
```{r plot_random_pairs}
threshold <- quantile(apply(dupls, 1, function(x) max(x)), 1 - obs_accel/obs_all)
# we'll use the same limits to make all the plots below more comparable
plim <- c(r_ave - 3.8 * r_sd, r_ave + 3.8 * r_sd)
plot(dupls, xlab = "Gene A", ylab = "Gene B", pch = 20, col = "blue", asp = 1, xlim = plim, ylim = plim)
accel <- dupls[(dupls$a >= threshold) | (dupls$b >= threshold), ]
abline(h = threshold, lwd = 2, lty = "dotted")
abline(v = threshold, lwd = 2, lty = "dotted")
#
# calculate the numbers and put those on the plot
both <- sum((accel$a >= threshold) & (accel$b >= threshold))
b_accel <- sum((accel$a < threshold) & (accel$b >= threshold))
a_accel <- sum((accel$a >= threshold) & (accel$b < threshold))
text(x = c(1.05*plim[1], 0.95*plim[2], 0.95*plim[2]), y = c(0.95*plim[2], 0.95*plim[2], 1.05*plim[1]), c(b_accel, both, a_accel), cex = 1.75)
```

### Correlated 

There are `r obs_all` pairs.  Each pair has an underlying characteristic rate, which we assign here at random. 
```{r family-specific_rates}
rates <- rnorm(obs_all, r_ave, r_sd)
```

Now we will simulate some diversity in observed rates for members of a pair, either due to sampling, or to divergence in underlying rate between the members of the pair. 

Here I am simulating 16 values for the standard deviation from the underlying rate.  
```{r random_correlated_pairs}
# add paired values that differ by mean = 0, sd
sds <- 16:1
results <- data.frame(sd = sds, single = NA, both = NA, R2 = NA)

# set up for multi-plot
d <- ceiling(sqrt(length(sds)))
oldpar <- par(no.readonly=TRUE)
par(mfrow=c(d, d), mar=c(0.25, 0.25, 0.5, 0.25), omi=c(0.25, 0, 0.25, 0))

for (sd in  sds) {
  dupls <- data.frame(a = rates + rnorm(obs_all, 0, sd), b = rates + rnorm(obs_all, 0, sd))
    # find the threshold that gives the right number of "accelerated" pairs
  threshold <- quantile(apply(dupls, 1, function(x) max(x)), 1 - obs_accel/obs_all)
  accel <- dupls[(dupls$a >= threshold) | (dupls$b >= threshold), ]
  # compare 
  both <- sum((accel$a >= threshold) & (accel$b >= threshold))
  single <- dim(accel)[1] - both
  results$both[results$sd == sd] <- both
  results$single[results$sd == sd] <- single
  results$R2[results$sd == sd] <- summary(lm(dupls$a ~ dupls$b))$r.squared
  
  # plot this  
  plot(dupls, axes = FALSE, frame.plot = TRUE, pch = 20, col = "blue", asp = 1, xlim = plim, ylim = plim, cex = 0.25)
  abline(h = threshold, lwd = 2, lty = "dotted")
  abline(v = threshold, lwd = 2, lty = "dotted")
# calculate the numbers and put those on the plot
  b_accel <- sum((accel$a < threshold) & (accel$b >= threshold))
  a_accel <- sum((accel$a >= threshold) & (accel$b < threshold))
  text(x = c(1.05*plim[1], 0.95*plim[2], 0.95*plim[2]), y = c(0.95*plim[2], 0.95*plim[2], 1.05*plim[1]), c(b_accel, both, a_accel), cex = 1, col = "red")
}
mtext("A range of correlations between members of a duplicate pair", outer = TRUE, cex = 0.8)
par(oldpar)
``` 

## The KBL model? 

A visualization of what Kellis seems to be thinking. 
```{r diagram_of_kbl_model}
plot(c(0,1), c(0,1), pch = "", xlab = "right member of pair", ylab = "left member of pair", asp = 1)
threshold <- 0.8
abline(h = threshold, lwd = 2, col = "grey", lty = "dotted")
abline(v = threshold, lwd = 2, col = "grey", lty = "dotted")
symbols(0.5, 0.9, circles = 0.05, inches = FALSE, add = TRUE, bg = "orange")
symbols(0.9, 0.5, circles = 0.05, inches = FALSE, add = TRUE, bg = "orange")
symbols(0.9, 0.9, circles = 0.025, inches = FALSE, add = TRUE, bg = "grey")
arrows(0.5, 0.5, 0.9, 0.5, lwd = 4, col = "red")
arrows(0.5, 0.5, 0.5, 0.9, lwd = 4, col = "red")
arrows(0.5, 0.5, 0.9, 0.9, lwd = 2, col = "black")
symbols(0.5, 0.5, circles = 0.15, inches = FALSE, add = TRUE, bg = "black")
text(0.5, 0.1, "not accelerated")
```

A binomial test comparing the number of asymmetric to symmetric accelerated pairs, consistent with the above model (except that a statistical test implies something other than the deterministic world imagined above).  
```{r binomial_test}
binom.test(c(obs_accel - obs_both, obs_both), p = 0.5, conf.level = 0.95)
```

